Dark matter web in the Universe

### 2018-06-11 Draft/Chapter1/intro.tex
thesis update

This thesis is a comprehensive collection of various cosmic structure analyses in the context of Lagrangian submanifold -- from topological studies of voids, to environmental investigation of dark matter haloes. In Chapter \ref{chapter2}, we give a brief overview of topics related to rest of the thesis -- including Zeldovich Approximation. Lagrangian submanifold and multistream fields. Chapter \ref{chapter3} approaches to quantify thresholds based on multistreams with the visible dark matter structures and Chapter \ref{chapter3b} compares this approach with a variety of current structure finding algorithms. The next two chapters deal with deeper understading of topological and geometrical feautres detected in multistream field (in Chapter \ref{chapter4}) and a new approach of detecting dark matter haloes that is independent of heuristic parameters (in Chapter \ref{chapter5}). Concluding Chapter \ref{chapter6} highlights future avenue of similar studies on the Lagrangian sub-manifold -- including detection of caustic surfaces and halo substructure analysis using flip-flop fields.

>>>>

This thesis is a comprehensive collection of various cosmic structure analyses in the context of Lagrangian submanifold -- from topological studies of voids, to environmental investigation of dark matter haloes. In Chapter \ref{chapter2}, we give a brief overview of topics related to rest of the thesis -- including Zeldovich Approximation. Lagrangian submanifold and multistream fields. Chapter \ref{chapter3} approaches to quantify thresholds based on multistreams with the cosmic web structures. The next two chapters deal with deeper understading of topological and geometrical feautres detected in multistream field (in Chapter \ref{chapter4}) and a new approach of detecting dark matter haloes that is independent of heuristic parameters (in Chapter \ref{chapter5}). Comparison of a variety of current cosmic structure finding algorithms are compared with our multistream-based methodology in Chapter \ref{chapter3b}. The concluding Chapter \ref{chapter6} highlights future avenue of similar studies on the Lagrangian sub-manifold -- including detection of caustic surfaces and halo substructure analysis using flip-flop fields.

### 2018-06-11 Draft/Chapter3/Paper2015.tex
thesis update

Using the full six-dimensional information allows one to generate new fields which  provide 
additional useful information about the evolution and morphology of the  web. 
One of them is a multi-stream field in Eulerian space, which will be the focus of this paper. 
Another example is the flip-flop field in Lagrangian space. In cosmological context it was first used in the ZA.
\citet{Vogelsberger2011} used it in a study of multi-stream structure of galaxy size haloes.
\citet{Shandarin2014a} applied it for identifying subhaloes  in dark matter haloes.
A similar although somewhat simplistic realization of this idea has been revealed in the ORIGAMI method
used for the analysis of the web ( \citealt{Falck2012},  \citealt{Falck2015} ).
 Although these fields cannot be used directly on observational data because the full phase-space information
 is not available, they provide much deeper insight into non-linear clustering of collision-less dark matter
and reveal new features of the web.

@@

Phase space tessellation considers the dynamics of the particles similar to that of a standard N-body code. 
However the particles are nodes of the tessellation, and are just massless tracers of the flow.
Assuming that the uniform state is modelled by a simple rectangular grid, the particles are the nodes of the grid.
Each elementary cube of the grid is tessellated by five tetrahedra (\citealt{Shandarin2012}\footnote{For the description of an alternative type of the tessellation see \citet{Abel2012}.}) of which the vertices are the
vertices of the cube. Mass is assumed to be uniformly distributed within each tetrahedron and the tessellation remain
intact at all times. The tetrahedra of the tessellation change their shapes and volumes, the latter are used for
computing the densities of the tetrahedra. Despite the complicated deformations experienced by the three-dimensional
sub-manifold tessellated by the tetrahedra, it remains continuous. Projected on three-dimensional configuration space, the tetrahedra
may form complicated structures. The number of streams at a chosen point {\bf x} is simply the number of tetrahedra that contain the point. The diagnostic points are computationally convenient to choose on a regular grid which can be significantly finer than the original grid in Lagrangian space. The ratio of separation of particles on the initial unperturbed grid to the separation distance of points in the diagnostic grid  $l_{\rm part}/l_{\rm dg} $ will be referred to as the refinement factor in the rest of the paper.

@@

Since each multi-stream region is surrounded by lower number of streams, the walls sandwich filaments within themselves (Fig. \ref{fig:4view}). The filaments are embedded with haloes at various intersections. These high-streaming haloes are completely covered by relatively low-stream filaments and hence surrounded by walls too. This result differs considerably from the several void finder methods, which find existence of haloes within  void regions (See \citealt{Colberg2008} and references therein). By our classification, we distinguish configuration space of the simulation box as void and non-void or web regions. Further, we have made an attempt to classify the web into walls, filaments and haloes based on multi-stream thresholds. This classification based on number of stream threshold provides only a very crude description of visual impression from the richness, complexity and fundamentally multi-scale character of the web. The heuristic numbers we use in this paper are by no means universal, but may provide limited use in the discussion of these particular simulations.

@@

\section{Summary}
In this paper, we explore for the first time, the multi-stream environment of dark matter haloes in cosmological N-body simulations.
%We study the multi-stream field computed in N-body simulations in $\Lambda$CDM cosmology at $z=0$.
The visualization and quantitative characterization of generic non-linear fields in three-dimensional space represent a serious
challenge from both conceptual and computational points of view. The complexity of the problem requires diverse tools
for analysing the results of cosmological simulations as well as galaxy catalogues.

@@

In this study we introduced an empirical statistical criteria which very crudely distinguish wall, filament and haloes. %for identifyinh filaments and walls.
We have found empirically that  in  the studied simulation the transition from wall points to filament points takes place approximately at $ 5 \le n_{str} \lesssim 15$. Using the virial over-density threshold of 200 in Eq.~\ref{eq:den}, we have also estimated that the haloes correspond to the regions with $n_{str} \gtrsim 90$.
Thus, the transition from filament to haloes takes place in the range $ 17 \le n_{str} \lesssim 90$.
The above critical values for transition from walls to filaments and from filaments to haloes were shown to be approximately correct
for  the simulation with $L/N = 0.78 h^{-1}$ Mpc. This technique can be also applied to simulations with different $L/N$ ratio and multi-stream grid of different refinement factors but the classification based on the threshold applied locally will remain only a very crude estimator. A more sophisticated morphological analysis will require non-local geometric and topological methods, which is beyond the scope of this paper.

>>>>

Using the full six-dimensional information allows one to generate new fields which  provide 
additional useful information about the evolution and morphology of the  web. 
One of them is a multi-stream field in Eulerian space, which will be the focus of this chapter. 
Another example is the flip-flop field in Lagrangian space. In cosmological context it was first used in the ZA.
\citet{Vogelsberger2011} used it in a study of multi-stream structure of galaxy size haloes.
\citet{Shandarin2014a} applied it for identifying subhaloes  in dark matter haloes.
A similar although somewhat simplistic realization of this idea has been revealed in the ORIGAMI method
used for the analysis of the web ( \citealt{Falck2012},  \citealt{Falck2015} ).
 Although these fields cannot be used directly on observational data because the full phase-space information
 is not available, they provide much deeper insight into non-linear clustering of collision-less dark matter
and reveal new features of the web.

@@

Phase space tessellation considers the dynamics of the particles similar to that of a standard N-body code. 
However the particles are nodes of the tessellation, and are just massless tracers of the flow.
Assuming that the uniform state is modelled by a simple rectangular grid, the particles are the nodes of the grid.
Each elementary cube of the grid is tessellated by five tetrahedra (\citealt{Shandarin2012}\footnote{For the description of an alternative type of the tessellation see \citet{Abel2012}.}) of which the vertices are the
vertices of the cube. Mass is assumed to be uniformly distributed within each tetrahedron and the tessellation remain
intact at all times. The tetrahedra of the tessellation change their shapes and volumes, the latter are used for
computing the densities of the tetrahedra. Despite the complicated deformations experienced by the three-dimensional
sub-manifold tessellated by the tetrahedra, it remains continuous. Projected on three-dimensional configuration space, the tetrahedra
may form complicated structures. The number of streams at a chosen point {\bf x} is simply the number of tetrahedra that contain the point. The diagnostic points are computationally convenient to choose on a regular grid which can be significantly finer than the original grid in Lagrangian space. The ratio of separation of particles on the initial unperturbed grid to the separation distance of points in the diagnostic grid  $l_{\rm part}/l_{\rm dg} $ will be referred to as the refinement factor in the rest of the chapter.

@@

Since each multi-stream region is surrounded by lower number of streams, the walls sandwich filaments within themselves (Fig. \ref{fig:4view}). The filaments are embedded with haloes at various intersections. These high-streaming haloes are completely covered by relatively low-stream filaments and hence surrounded by walls too. This result differs considerably from the several void finder methods, which find existence of haloes within  void regions (See \citealt{Colberg2008} and references therein). By our classification, we distinguish configuration space of the simulation box as void and non-void or web regions. Further, we have made an attempt to classify the web into walls, filaments and haloes based on multi-stream thresholds. This classification based on number of stream threshold provides only a very crude description of visual impression from the richness, complexity and fundamentally multi-scale character of the web. The heuristic numbers we use in this chapter are by no means universal, but may provide limited use in the discussion of these particular simulations.

@@

\section{Summary}
In this chapter, we explore for the first time, the multi-stream environment of dark matter haloes in cosmological N-body simulations.
%We study the multi-stream field computed in N-body simulations in $\Lambda$CDM cosmology at $z=0$.
The visualization and quantitative characterization of generic non-linear fields in three-dimensional space represent a serious
challenge from both conceptual and computational points of view. The complexity of the problem requires diverse tools
for analysing the results of cosmological simulations as well as galaxy catalogues.

@@

In this study we introduced an empirical statistical criteria which very crudely distinguish wall, filament and haloes. %for identifyinh filaments and walls.
We have found empirically that  in  the studied simulation the transition from wall points to filament points takes place approximately at $ 5 \le n_{str} \lesssim 15$. Using the virial over-density threshold of 200 in Eq.~\ref{eq:den}, we have also estimated that the haloes correspond to the regions with $n_{str} \gtrsim 90$.
Thus, the transition from filament to haloes takes place in the range $ 17 \le n_{str} \lesssim 90$.
The above critical values for transition from walls to filaments and from filaments to haloes were shown to be approximately correct
for  the simulation with $L/N = 0.78 h^{-1}$ Mpc. This technique can be also applied to simulations with different $L/N$ ratio and multi-stream grid of different refinement factors but the classification based on the threshold applied locally will remain only a very crude estimator. A more sophisticated morphological analysis will require non-local geometric and topological methods, which is discussed in Chapters \ref{chapter4} and \ref{chapter5}.

### 2018-06-11 Draft/Chapter4/Paper2017a_Nobmath.tex
thesis update

Heuristic analysis of the multistream field has been done in Chapter \ref{chapter3} and parts of Chapter \ref{chapter3b}. In this chapter, we analyse the dark matter structures based on their phyiscal shapes and connectivity. Extensive studies of geometrical and topological properties in cosmic matter density field have been done in the past. Here we extend some of these studies into the Lagrangian sub-manifold regime -- taking advantage of certain unique properties of the multistream field supplement our understanding of dark matter structures. The analysis and text in this chapter overlaps with \cite{Ramachandra2017}.

@@

There are extensive number of topological indicators in the context of density fields or spatial co-ordinates - such as alpha shapes, Betti numbers, genus statistics. Although a comparitive study of these topolgical measures in multistream fields may be interesing, it is not the intent of this paper. In this study, we only investigate percolation transitions in excursion sets of multistreams as a preliminary analysis of topological connectivities. Excursion sets in density fields are shown to have quick percolation transitions \citep{Shandarin2010b} and a similar trend in multistream field is investigated here.

>>>>

Heuristic analysis of the multistream field has been done in Chapter \ref{chapter3} where the dark matter structures were classified based on local multistream variational approach. In this chapter, we analyse the dark matter structures based on their phyiscal shapes and connectivity. Extensive studies of geometrical and topological properties in cosmic matter density field have been done in the past. Here we extend some of these studies into the Lagrangian sub-manifold regime -- taking advantage of certain unique properties of the multistream field supplement our understanding of dark matter structures. The analysis and text in this chapter overlaps with \cite{Ramachandra2017}.

@@

There are extensive number of topological indicators in the context of density fields or spatial co-ordinates - such as alpha shapes, Betti numbers, genus statistics. Although a comparitive study of these topolgical measures in multistream fields may be interesing, it is not the intent of this chapter (See Chapter \ref{chapter3b} for a detailed comparison of various structure finding algorithms). In this study, we only investigate percolation transitions in excursion sets of multistreams as a preliminary analysis of topological connectivities. Excursion sets in density fields are shown to have quick percolation transitions \citep{Shandarin2010b} and a similar trend in multistream field is investigated here.

### 2018-06-11 Draft/Chapter5/Paper2017b_Nobmath.tex
thesis update

Identifying the haloes from the distribution of particles in N-body simulations is one of the problems attracting both considerable interest and efforts. In this chapter, we propose a novel framework for detecting potential dark matter haloes using the field unique for dark matter  -- multistream field. The current status of halo finding techniques are summarized in Section \ref{sec:intro}. A review of the DM particle clustering in a one-dimensional dimensional universe is made in Section \ref{sec:HaloFormation}, and the concept of multistream field is extended to higher dimensions. The multistream field is computed on the cosmological simulations described in Section \ref{sec:ch5_simulation}. The halo identification framework in this field is described in Section \ref{sec:haloDetection}. This algorithm isolates convex regions of the multistream field using Hessian eigenvalues, each enclosing a local multistream maximum. Without employing any non-local thresholds that several halo finders generally use, these convex multistream regions are identified as potential halo sites. We also illustrate the significance of multistream refinement and softening scales in finding subhaloes. However, this paper does not focus on adaptive multi-scale analyses for substructure studies. A few properties of the multistream haloes are discussed in Section \ref{sub:HaloProperties}, and comparison of these haloes with AHF and FOF algorithms is done in Section \ref{sub:compareHalo}. We also discuss the spatial distribution of the dark matter haloes along the percolating web structure.

@@

The emphasis of this paper is to demonstrate the use of multistream field in identifying potential dark matter haloes, and not a full statistical analysis of halo properties. For this purpose, we have run simulations at two different mass resolutions (number of particles $N_p = 128^3$ and $256^3$, and respective mass of particles, $m_p = 3.65 \times 10^{10} h^{-1} M_{\odot}$ and $ 4.57 \times 10^{9} h^{-1} M_{\odot}$), with the same periodic side length $L = 100 h^{-1} Mpc$. The gravitational softening length $\epsilon = 20 h^{-1} kpc$ and $10 h^{-1} kpc$ for low and high resolution simulation respectively. The initial conditions are generated by {MUSIC} \citep{Hahn2011a} with the transfer function from \cite{Eisenstein1998a} at a redshift of $z_{ini}= 80$. The $\Lambda$CDM cosmological simulation is run using {GADGET-2} (\citealt{Springel2005c} and \citealt{Springel2001}) is similar to the ones used in \cite{Ramachandra2017}. The cosmological parameters used in the simulation are $\Omega_{m}= 0.276$, $\Omega_{\Lambda}= 0.724$, the Hubble parameter, $h = 0.703$, the power spectrum normalization, $\sigma_8 = 0.811$ and the spectral index $n_s= 0.961$.

@@

The requirement that each halo should have closed convex surfaces with a multistream maximum inside may identify sub-haloes in large haloes but is too demanding because a halo with sub-haloes must have saddle points in the $n_{str}$ field. This may explain the shortage of massive haloes shown in \autoref{fig:hmf}. Applying a filter for smoothing the $n_{str}$ field increases the number of massive haloes but reduces the number of low mass haloes. Although our present method does not currently perform an analysis simultaneously on multiple smoothing scales, such approaches done in density, lognormal density, tidal, velocity divergence or velocity shear fields (see MMF by \citealt{Aragon-Calvo2007} and NEXUS+ by \citealt{Cautun2013}) have shown interesting multi-scale features of the cosmic web. Applying a more sophisticated procedure for linking $\lambda_3$-sub-haloes in a more massive halo will be done in the follow-up paper.

>>>>

Identifying the haloes from the distribution of particles in N-body simulations is one of the problems attracting both considerable interest and efforts. In this chapter, we propose a novel framework for detecting potential dark matter haloes using the field unique for dark matter  -- multistream field. The current status of halo finding techniques are summarized in Section \ref{sec:intro}. A review of the DM particle clustering in a one-dimensional dimensional universe is made in Section \ref{sec:HaloFormation}, and the concept of multistream field is extended to higher dimensions. The multistream field is computed on the cosmological simulations described in Section \ref{sec:ch5_simulation}. The halo identification framework in this field is described in Section \ref{sec:haloDetection}. This algorithm isolates convex regions of the multistream field using Hessian eigenvalues, each enclosing a local multistream maximum. Without employing any non-local thresholds that several halo finders generally use, these convex multistream regions are identified as potential halo sites. We also illustrate the significance of multistream refinement and softening scales in finding subhaloes. However, this chapter does not focus on adaptive multi-scale analyses for substructure studies. A few properties of the multistream haloes are discussed in Section \ref{sub:HaloProperties}, and comparison of these haloes with AHF and FOF algorithms is done in Section \ref{sub:compareHalo}. We also discuss the spatial distribution of the dark matter haloes along the percolating web structure.

@@

The emphasis of this chapter is to demonstrate the use of multistream field in identifying potential dark matter haloes, and not a full statistical analysis of halo properties. For this purpose, we have run simulations at two different mass resolutions (number of particles $N_p = 128^3$ and $256^3$, and respective mass of particles, $m_p = 3.65 \times 10^{10} h^{-1} M_{\odot}$ and $ 4.57 \times 10^{9} h^{-1} M_{\odot}$), with the same periodic side length $L = 100 h^{-1} Mpc$. The gravitational softening length $\epsilon = 20 h^{-1} kpc$ and $10 h^{-1} kpc$ for low and high resolution simulation respectively. The initial conditions are generated by {MUSIC} \citep{Hahn2011a} with the transfer function from \cite{Eisenstein1998a} at a redshift of $z_{ini}= 80$. The $\Lambda$CDM cosmological simulation is run using {GADGET-2} (\citealt{Springel2005c} and \citealt{Springel2001}) is similar to the ones used in \cite{Ramachandra2017}. The cosmological parameters used in the simulation are $\Omega_{m}= 0.276$, $\Omega_{\Lambda}= 0.724$, the Hubble parameter, $h = 0.703$, the power spectrum normalization, $\sigma_8 = 0.811$ and the spectral index $n_s= 0.961$.

@@

The requirement that each halo should have closed convex surfaces with a multistream maximum inside may identify sub-haloes in large haloes but is too demanding because a halo with sub-haloes must have saddle points in the $n_{str}$ field. This may explain the shortage of massive haloes shown in \autoref{fig:hmf}. Applying a filter for smoothing the $n_{str}$ field increases the number of massive haloes but reduces the number of low mass haloes. Although our present method does not currently perform an analysis simultaneously on multiple smoothing scales, such approaches done in density, lognormal density, tidal, velocity divergence or velocity shear fields (see MMF by \citealt{Aragon-Calvo2007} and NEXUS+ by \citealt{Cautun2013}) have shown interesting multi-scale features of the cosmic web. 
% Applying a more sophisticated procedure for linking $\lambda_3$-sub-haloes in a more massive halo will be done in the follow-up paper.

### 2018-06-11 Draft/Chapter6/concl.tex
thesis update

\chapter{Conclusions and Future}

In this Chapter, we present our general conclusions. We discuss the limitations inherent
in our model, and speculate upon how we may address these weaknesses in the future.
We, furthermore, speculate upon the possible applications of our model to real plasmas.
We, finally, consider the physical questions still remaining – concluding, ultimately, that a
considerable amount of work is still needed to answer these questions definitively.

@@

\subsection{Flip-Flop}
\subsection{Caustics}
\subsection{Simulation}

@@

>>>>

\chapter{General Conclusions}\label{chapter6}

Formation and evolution velocity multistreams in the dark matter Universe plays a vital role in the formation of large scale structures in the Universe. In this thesis, we have analysed this multi-stream fields in cosmological context -- from dark matter halo environments to connectivities in the single streaming voids. It has been demonstrated that the studies of the Lagrangian submanifold not only provides a novel way of looking at the cosmos, but also boosts our current knowledge of the dark matter Universe.

This final chapter encapsulates signifcant findings of our work. In addition, we also highlight a few interesting avenues for future research using the Lagrangian submanifold.

\section{Summary of major results}

The full dynamical state of  dark matter can be described as a three-dimensional sub-manifold
in six-dimensional phase space - the dark matter sheet. In our study we use a Lagrangian sub-manifold ${\bf x} = {\bf x}({\bf q},t)$ 
(where {\bf x} and {\bf q} are co-moving Eulerian and Lagrangian coordinates respectively), which 
is dynamically  equivalent to the dark matter sheet but is more convenient for numerical analysis. Projecting the Lagrangian sub-manifold at each point in the configuration space defines the multistream field  $n_{str}(\mathbf{x}, z)$. This integer-valued field is equivalent to counting the number of foldings in the submanifold and we numerically calculate by a Lagrangian tessellation scheme. The resulting multistream field is dynamical, positively correlated with matter density, and supplements our knowledge of spatial clustering. Our major results can be summarized as follows.

\subsection{Heirarchial structure of the dark matter web}

The first study of the multi-stream environment of dark matter haloes in cosmological N-body simulations in the $\Lambda$CDM cosmology reported in Chapter \ref{chapter3} (first shown in \cite{Ramachandra2015}) shows that at the resolution of the simulation i.e. without additional smoothing, the cosmic web represents a hierarchical structure: each halo is embedded in the filamentary framework of the web predominantly  at the filament crossings, and each filament is embedded in the wall like fabric of the web at the wall crossings.

Voids are uniquely defined by the local condition requiring to be a single-stream flow region. Within the dark matter web (regions with $n_{str} > 1$), The walls are regions where number of streams is equal to three or a few. The number of streams in the neighbouring filaments is higher than in the neighbouring walls. Finally, each halo or sub-halo is a local peak in the $n_{str}$ field. We demonstrated in Chapter \ref{chapter3} that the shells of streams around haloes are quite thin and the closest void region is typically within one and a half FOF radius from the center of the halo.

\subsection{Multistream view of the dark matter haloes}

In our simulation reported in Chapter \ref{chapter3}, simple virial density criteria corresponds to a global threshold of $n_{str} \geq 90$ for DM haloes. This shows a reasonably good correspondence with structure finders. A more sophisticated halo finding formulation is done in Chapter \ref{chapter5} assuming that the virialized haloes have convex boundaries. Closed and convex regions of the multistream field are hence isolated by imposing a positivity condition on all three eigenvalues of the Hessian estimated on the smoothed multistream field. In a single-scale analysis of high multistream field resolution and low softening length, the halo substructures with local multistream maxima are isolated as individual halo sites. This halo finding technique is free of any heuristic parameters, and maybe extended to delineate filaments, walls, voids, as well as halo sub-structures. Agreement of halo summary statistics (like the halo mass function) of the multistream haloes with traditional halo finders (Friends-of-Friends and Spherical Overdensity based techniques) is shown \cite{Ramachandra2017b}. As opposed to several other halo finders, this geometrical approach accounts for non-spherical boundaries, does not find any halo candidates in voids, and minimizes the need for post-processing in the production of halo catalogues for dark matter simulations.

\subsection{Topological transitions in the cosmic web}

Topological connections in the single-streaming voids and multistreaming filaments and walls reveal a cosmic web structure different from traditional mass density fields, as shown in Chapter \ref{chapter4}. Single streaming voids occupy over 90 per cent of the dark matter Universe. A single void structure not only percolates the multistream field in all the directions, but also spans over 99 per cent of all the single-streaming regions. Sub-grid analyses on scales smaller than simulation resolution reveal tiny pockets of voids that are isolated by membranes of the structure.

Rapid topological transitions from are seen in matter density fields as well as multistream fields, i.e., at a given threshold (matter density or multistream), the filling fraction of the largest structure in the Universe dominates the excursion volume fraction. For the multistreaming excursion sets, the percolating structure is significantly thinner than the filaments in over-density excursion approach. We also demonstrated that the matter density field is more fragmented than the multistream field.

@@

Although the scope of this thesis has primarily been multistream field and its application to cosmic structure formation, it is necessary to highlight other important physical insights derived from the Lagrangian submanifold. For instance, tracing the Lagrangian sub-manifold also provides rich insights into halo collapse modelling \cite{Neyrinck2016}. Recently, there are attempts to improve N-body simulations (see \cite{Hahn2013}, \cite{Angulo2013a}, \cite{Angulo2013b}, \cite{Sousbie2015} and \cite{Hahn2016a}) by solving the Vlasav-Poisson equation using tessellations in the Lagrangian sub-manifold. Galaxy evolution and star formation in the context of multi streaming phenomenon are studied by \cite{Aragon-Calvo2016}. Problems of halo-core tracking, dynamical analysis of halo merger histories, halo and subhalo disruption events and mass accretion of filaments and haloes may be revisited using Lagrangian submanifold.

This section reviews two of the potential avenues for interesting application of Lagrangian submanifold. One may track the {\it Flip-Flop} field -- to understand the rich and comples substructures within haloes. On the other hand, Lagrangian submanifold may also be used to identify caustic surfaces -- measure-zero surfaces in the DM Universe that have formally infinite density.

\subsection{Flip-Flop analyses in Lagrangian field}

The integer value of Flip-flop in Lagrangain field, $n_{ff}(\mathbf{q})$ is increased by every swap of the Eulerian coordinates of the two neighbouring DM particles. This number field is estimated in three dimensions by computing the Jacobian  $J(\mathbf{q}, t) = |\frac{\partial\mathbf{x}}{\partial\mathbf{q}}|$  on each particle at each time step. If the sign of the Jacobian changes, the number of flip-flops for the corresponding particles is increased by one. Details of calculation of flip-flop field and it's significance in dynamical history of substructure formation is explained in \citep{Shandarin2016}, and summarized in Section \ref{sec:ZA}. This Lagrangian field $n_{ff}(\mathbf{q})$ for a cosmological simulation with $128^3$ DM particles and side length of $100 h^{-1} Mpc$ is shown in the left panel of \autoref{fig:ffq}. Halo environment analysis similar to one in Chapter \ref{chapter5} may allow us to identify and track the formation of DM haloes directly from the Lagrangian space. Alternatively, this also enables back-tracking the DM particles in halos cores and substructures all the way to the initial configurations in the simulation.

\begin{figure}
\begin{minipage}[t]{.99\linewidth}
 \centering\includegraphics[width=5.7cm]{Chapter2/Plots/ffq.pdf} 
\centering\includegraphics[width=5.cm]{Chapter2/Plots/ff0_slice.pdf} 
 \centering\includegraphics[width=5.cm]{Chapter2/Plots/ff1_slice.pdf} 
\end{minipage}\hfill
\caption{Left panel: Lagrangian slice of the Flip-Flop field $n_{ff}(\mathbf{q})$ in a cosmological simulation with $L = 100 h^{-1} Mpc$. The values of $n_{ff}$ varies from $0$ to $15$. Center panel: Eulerian picture, only the particles with $n_{ff}(\mathbf{q}) = 0$. are shown. Red regions show regions with $n_{str}(\mathbf{x}) > 1$. Right panel: Eulerian picture, only the particles with $n_{ff}(\mathbf{q}) > 0$. are shown. Red regions show regions with $n_{str}(\mathbf{x}) > 1$.}
\label{fig:ffq}
\end{figure}

\begin{figure}
\begin{minipage}[t]{.99\linewidth}
%  \centering\includegraphics[width=5.cm]{Chapter2/Plots/PercPlot_ns.pdf} 
 \centering\includegraphics[width=8.cm]{Chapter2/Plots/PercPlot_ff.pdf} 
  \centering\includegraphics[width=8.cm]{Chapter2/Plots/PercPlot_ff_ns.pdf}  
\end{minipage}\hfill
\caption{Left panel: Percolation transition for the Flip-Flop, $n_{ff}(\mathbf{q})$. Shows the excursion set fraction $f_{ES}$ and the filling fraction $f_1$ for various flip-flop thresholds. Right panel: Percolation transition for the Flip-Flop field, $n_{ff}(\mathbf{q})$ and multi-stream field $n_{str}(\mathbf{x})$.}
\label{fig:percplot}
\end{figure}

Since the $n_{ff}$ field is defined on particles, mapping the field to Eulerian field is straigtforward. Correspondence between the multistream field defined in Eulerian space $n_{str}(\mathbf{x})$ is shown in the middle and right panels of \autoref{fig:ffq}. Most particles that have not undergone a flip-flop are in the single-streaming region. In the multistreaming region, however, consists of particles with $n_{ff} = 0$ and $n_{ff} > 0$. On the other hand, $n_{ff} > 0$ particles are not found in single-streaming voids.

Percolation analysis (similar to the multistream analysis in Chapter \ref{chapter4}) may also be done using the Flip-Flop in the Lagrangian field. \autoref{fig:percplot} shows topological transitions similar to ones in multistream and matter density fields shown in \autoref{fig:PercTh} and \autoref{fig:HaloFil}, except here we explore the connectivities in the Lagrangian space. This could reveal topological connections in the initial positions, and reveal whether or not the connections remain intact throughout the evolution of nonlinear gravitational collapse.

% \begin{figure}
% \begin{minipage}[t]{.99\linewidth}
%  \centering\includegraphics[width=8.cm]{Chapter2/Plots/PercPlot_ff_ns.pdf}  
% \end{minipage}\hfill
% \caption{Percolation transition for the Flip-Flop, $n_{ff}(\mathbf{q})$ and multi-stream field $n_{str}(\mathbf{x})$.}
% \label{fig:perctransition}
% \end{figure}

\subsection{Caustics formation}

Existence of singularities is a distingushing feature of collisionless DM collapse from that of baryonic matter. Caustic formation in the context of the ZA \cite{Zeldovich1970} was discussed earlier in Section \ref{sec:ZA} (also see \cite{Shandarin1989} for a detailed study of cosmic singularities) . Theoretical characterization of caustic surfaces in a two dimensional gravitational collapse was studied by \citep{Arnold1982}. However, such analytical treatment is complicated in the three-dimensional case (See for example, the discussions in \citealt{Hidding2014}, \citealt{Feldbrugge2018}). Numerically, one may find the particles belonging to caustic surfaces by following the determinants of the volume elements in the Lagrangian submanifold. This result of such scheme was demonstrated in \autoref{fig:NstrCaust}.

% \begin{equation} 
% \label{eq:det}
% V = \frac{1}{3!}
% \begin{vmatrix}
% x_{1} & y_{1} & z_{1} & 1 \\ 
% x_{2} & y_{2} & z_{2} & 1 \\ 
% x_{3} & y_{3} & z_{3} & 1 \\ 
% x_{4} & y_{4} & z_{4} & 1 
% \end{vmatrix}
% \end{equation}

Matter density is formally infinite at the location of caustic surfaces, where dark matter sheet folds in phase-space. Being measure-zero structures, identification of caustics via matter density fields is usually restricted to fine-grained simulations. Caustics are clearly related to multistream or Flip-Flop fields as well. Multistream values are generally odd-valued - i.e., the $n_{str}(\mathbf{x}) = 1, 3, 5, 7$  and so on. Exception occur at Eulerian positions of caustic surfaces $x_c$, where are there are even number of streams. Unfortunately, these are not easily resolved since the structures of measure zero. However, multistream fields do resolve positions of caustic surfaces to the level of resolution of the field. In other words, any transition in $n_{str}(\mathbf{x})$ value signifies location of caustic surface between the gradient. On the other hand, Flip-Flop fields have a more straightforward relationship since they are defined on the particles. In one dimensional collapse, caustic locations are simply the particles where the value of $n_{ff}$ changes.

\begin{figure} 
\centering\includegraphics[width=9cm]{Chapter2/Plots/f1fES_caustic.pdf} 
\caption{Percolation in caustic particles with variation of $l_{max}$: Top panel shows the mass fraction of the largest isolated caustic surface $f_1$ and mass fraction of all particles in the caustic $f_{ES}$. Bottom panel shows the filling fraction $f_1/f_{ES}$. Between 3.5-4 $h^{-1} Mpc$ the caustic surfaces transition into distinct isolated surfaces -- i.e., forming turn-around boundaries of haloes. }
\label{fig:caustic_perc_lmax}
\end{figure}

\begin{figure*} 
\centering\includegraphics[height=15cm]{Chapter2/Plots/caustic_connectivity.png} 
\caption{$L=20$ Mpc, $N_p = 32^3$. Small simulation, Caustic are filtered by longest side length of the triangle: Top left: $ l_{max} \geq 0$, Top right: $ l_{max} \geq 1$, Bottom left: $ l_{max} \geq 2$, Bottom right: $ l_{max} \geq 3$. }
\label{fig:caustic_lmax}
\end{figure*}

Despite the theoretical understanding mentioned above, delineating caustic surfaces at various levels of gravitational collapse has been proven to be difficult. Instead, a combination of geometric and topological methods may also enable us to trace the caustic formation. The DM particles on the caustics are on curved two dimensional surfaces, forming vertices of triangles that constitute the {\it tiles} of the surface. These triangles are of various shapes and size. Since the mean separation of particles is smaller than elsewhere in the cosmic web, the triangles are smaller on an average. Using the largest side of the triangle $l_{max}$ as a proxy for length scale of the triangle, one may perform morphological analyses on the type of caustic surfaces. One such study is shown in \autoref{fig:caustic_perc_lmax}. Percolation transition (in an unstructured grid, as opposed to regular grid analyses for multistream or flip-flop fields) are used to differentiate the larger casutic surface (like the ones between voids and walls) and the smaller ones (like the concentric caustic shells around haloes). This transition in our small cosmological simulation of side length $L = 20 h^{-1} Mpc$ is seen at around $3.5-4 h^{-1} Mpc$.

Using the heuristic parameter of $l_{max}$, we are able to isolate caustic surfaces around specific structures like the haloes. The \autoref{fig:caustic_lmax} shows all the caustic surfaces in the top left panel, and the inner caustics filtered using $l_{max}$ thresholds in the subsequent panels. Another interesting aspect of this filtering scheme is the ability to isolate the caustic surfaces around the haloes as seen in the bottom right panel of \autoref{fig:caustic_lmax}.

Careful examination of the smaller caustic surfaces around the haloes may reveal concentric shells where the particles {\it turn back}. A spherical approximation of this, called the {\it splashback radius} \citep{More2015}. This has gained a considerable attention recently due to possible observational signatures of physical boundaries of dark matter haloes \citep{Chang2017}. Caustic finding scheme with appropriate filtering provides a general method to find these boundaries exactly. Moreover, mapping caustic surfaces from Eulerian space to Lagrangian space displays concentric surfaces, providing a dynamical insight into the splashback radii of dark matter haloes.

@@

Understanding the complex dynamical processes that shape the dark matter web has been the primary focus of this thesis. By developing a wide variety of topological, geometrical and statistical diagnostic algorithms, we have demonstrated the potential of the Lagrangian submanifold in unveiling properties of the cosmic structures. Despite these modern advancements, the Zel'dovich approximation has to be recognized as the basis for a dynamical theory for the evolution of the dark matter web and its components.

On the one hand, such a motivated theoretical framework may result in a systematic discriptor of the large scale dark matter structures in the Universe. On the other, perplexing complexities of halo substructures, caustic singularities and merger histories could be unmasked. This only demands extensive studies in the field of Physical cosmology.
